---
title: "Analysis of the climatic events that impacts health and economy of the EUA."
author: "Marco Pollo Almeida"
date: "2/11/2020"
---
        

## 1. Outline
This project involves exploring the U.S. National Oceanic and Atmospheric 
Administration's (NOAA) storm database. Storms and other severe weather 
events can cause both public health and economic problems for communities 
and municipalities. To prevent such outcomes to the extent possible is a 
key concern. The dataset presents the 1950 to 2011 period.

Based on the documentation of the dataset [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf), one investigates the main events which cause harm to Health (injuries and fatalities) and Economic Field (properties and crops).


## 2. Data Processing

### 2.1. Data Loading

Following steps: Downloading the dataset (raw data). Extract it from bzip2 file. Read and convert it into a "data.table" object.
```{r}
# Load packages that will be used
library(data.table)
library(tidyverse)

# Download and extract the file
URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(URL, destfile = paste0("~/", '/repdata%2Fdata%2FStormData.csv.bz2'))

# read the data
dt <- read.csv("~/repdata%2Fdata%2FStormData.csv.bz2")

# Transform to "data.table" object
dt <- as.data.table(dt)
```

### 2.2. Inspecting variables (name, type, class, etc.)

```{r}
str(dt)
```

### 2.3. Create new dataset from the raw data (subsetting).

Create new dataset based on the requirements of the project.
```{r results="hide"}
# Variables to remove
rmVarb <- colnames(dt[, !c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")])

# Removing variables
dt[, c(rmVarb) := NULL]

# Filtering data only with fatalities or injuries  
dt <- dt[(EVTYPE != "?" & (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP") ]
```

### 2.4. Transforming some characters according to documentation, in order to make possible the calculations.

Clean the PROPDMGEXP and CROPDMGEXP variables in order to calculate some results related to cost crop and properties.
```{r results="hide"}
# Change some characters to uppercase.
cols <- c("PROPDMGEXP", "CROPDMGEXP")
dt[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]
# Change alphanumeric exponents to numeric values (for PROPERTY variable).
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)
# Change alphanumeric exponents to numeric values.
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)
dt[, PROPDMGEXP := propDmgKey[as.character(dt[,PROPDMGEXP])]]
dt[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]
dt[, CROPDMGEXP := cropDmgKey[as.character(dt[,CROPDMGEXP])] ]
dt[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]
```

### 2.5. Creating Economic Cost Variables

```{r}
dt <- dt[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### 2.6. Creating Total Property and Crop Cost Variables

```{r}
dt2 <- dt[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
dt2 <- dt2[order(-Total_Cost), ]
dt2 <- dt2[1:10, ]
head(dt2, 5)
```

### 2.7. Creating Total Fatalities and Injuries Variables

```{r}
dt3 <- dt[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
dt3 <- dt3[order(-FATALITIES), ]
dt3 <- dt3[1:10, ]
head(dt3, 5)
```

## 3. Results

### 3.1. Events that are Most Harmful to Population Health

Melting data.table so that it is easier to put in bar graph format 
```{r }
dt4 <- melt(dt3, id.vars="EVTYPE", variable.name = "bad_thing")
head(dt4, 5)
```

```{r}
# Create chart
healthChart <- ggplot(dt4, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
healthChart = healthChart + geom_bar(stat="identity", aes(fill=bad_thing), position="dodge")
# Format y-axis scale and set y-axis label
healthChart = healthChart + ylab("Frequency Count") 
# Set x-axis label
healthChart = healthChart + xlab("Event Type") 
# Rotate x-axis tick labels 
healthChart = healthChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
healthChart = healthChart + ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))
healthChart
```

### 3.2. Events that have the Greatest Economic Consequences

Melting data.table so that it is easier to put in bar graph format 
```{r }
dt5 <- melt(dt2, id.vars="EVTYPE", variable.name = "Damage_Type")
head(dt5, 5)
```

```{r }
# Create chart
econChart <- ggplot(dt5, aes(x=reorder(EVTYPE, -value), y=value))
# Plot data as bar chart
econChart = econChart + geom_bar(stat="identity", aes(fill=Damage_Type), position="dodge")
# Format y-axis scale and set y-axis label
econChart = econChart + ylab("Cost (dollars)") 
# Set x-axis label
econChart = econChart + xlab("Event Type") 
# Rotate x-axis tick labels 
econChart = econChart + theme(axis.text.x = element_text(angle=45, hjust=1))
# Set chart title and center it
econChart = econChart + ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))
econChart
```


